---
title: "MorphoTools version 1.0.0 tutorial"
author: | 
  | Marek Šlenker \<marek.slenker@savba.sk\>
  | Plant Science and Biodiversity Center
  | Bratislava, Slovakia
  | https://github.com/MarekSlenker/MorphoTools
date: "April 26, 2020"
output:
  pdf_document: default
  html_document: default
vignette: >
  %\VignetteIndexEntry{MorphoTools_tutorial}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

\tableofcontents

## 1. Introduction

The R package MorphoTools provides useful tools for


## 2. Obtaining and installing MorphoTools package

The R console and base system can be obtained at <http://www.r-project.org/>. Once R is installed, MorphoTools can be installed and loaded by typing the following commands into the R console:


```{r eval=FALSE, include=T}
install.packages(MorphoTools)
library(MorphoTools)
```
```{r include=F}
library(MorphoTools)
```

If you quit and restart R, you will not have to re-install the package but you will need to load it again (using the library function as shown above).

## 3. Data import and control

### 3.1 Data import

As with any software, the first thing you want to do is import your raw data. Data can be imported from text files, import from Excel spreadsheets through the clipboard in Windows also usually works well. For this tutorial, go into the "extdata" directory of the MorphoTools package installation, and find a file called "centaurea.txt". To find the paths to the package locations run `path.package(package = "MorphoTools")`. Open this file in a text editor and inspect its contents. This file contains 25 morphological characters of three diploid species of the *Centaurea phrygia* complex, abbreviated "ph", "ps" and "st" and the putative hybrid of the last two species abbreviated as "hybr". For details run `?centaurea`.

As you find in centaurea.txt file, following structure of input data is required: 

* the first row contains variable names.
* the next rows contains individuals, single individual per row.
* the first three columns include unique identifiers for individuals, populations and taxa/groups, respectively. Columns have to be named as “ID”, “Population” and “Taxon”.
* fourth and next columns stores morphological characters.

Avoidi spaces and special characters for column names. Missing values have to be represented as empty cells or by the text `NA` (not quoted).

Read the centaurea.txt file using the `read.morphodata` function, and assign the dataset a name of your choice (centaurea in this example) by typing:

```{r eval = TRUE, echo = TRUE}
centaurea = read.morphodata(file = system.file("extdata", "centaurea.txt", package = "MorphoTools"),
                            dec = ".",
                            sep = "\t")
```

Alternatively, you can use build-in data running `data(centaurea)`. The dataset now exists as a `morphodata` object in R. The following commands display, some basic information about the dataset.

```{r echo = TRUE, eval=T}
summary(centaurea)
```

```{r echo = TRUE, eval=T}
head( samples(centaurea) , n = 60) # show only first 60 samles
```

```{r echo = TRUE, eval=T}
populations(centaurea)
```

```{r echo = TRUE, eval=T}
taxa(centaurea)
```


### 3.2 Data control

Row data can contain typos in decimal points, missing values, 
, 

pouzijeme boxploty
- identifikacia outlierov
- rozdiely medzi znakmi

boxplot.character
boxplot.all


descr. tabulka - less comprehensive vay, but can be used for reporting Descriptive statistics of measured morphological characters in your paper. 


descr.tax
descr.pop
descr.all


correlations - vyhodit korelovane znaky


cormat
cormat.signifTest



na.meanSubst



Remove items

delete.taxon
delete.population
delete.charecter


popul.otu


Now that you have your dataset completed, it is not a bad idea to save a copy of it using `export.res` function. 

export.res




## 4. clust



## 5. pca



## 6. cda










## How to cite MorphoTools

sssssssssssssssssssssssssssssssssssssss

Feel free to email me at marek.slenker@savba.sk with any questions, comments, or bug reports!
